---
title: "From Roots to Riches: Unraveling the Impact of Parental Education on a Child's Learning Odyssey and Future Finances"
author: 
  - Lexi Knight
  - Xavier Ibanez-Padron
thanks: "Code and data are available at: https://github.com/Ludosss/IncomeEducationStudy"
date: today
date-format: long
abstract: "This study investigates the empowering impact of parents' level of education on their child's academic success and as a result the child's household net income. Taking advantage of a large dataset from the European Social Survey and specifically looking at French data gathered in 2020, this study dives into the relationships of these variables to determine the correlation between parental education, the child's education as well as income levels. The headline result demonstrates a positive correspondence between years of parental education on a greater chance the child strives for higher education after high school and this is linked to an increase in income in the career path of their choice. The implications of these findings embellish the effect parents' education and outlook on learning on their child's academic and workplace success."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r} 
# Chunk 1

#| message: false
#| echo: false

### Workspace setup ###

# Install packages
# install.packages("tidyverse")

# Load packages
library(tidyverse)

# Read in cleaned data
cleaned_income_education_data <- read_csv(here::here("outputs", "data", "ESS10_Clean.csv"))

```

# Introduction
  A study done on tenth graders in the Mardan district of Pakistan showed that children with educated parents have notably greater academic success and go onto higher levels of education as compared to children with uneducated parents (Idris,2020). The authors explain that parents have a great deal of influence on their child's knowledge, vocabulary and values. Parents with higher education promote meaningful discussion at home, social skills, give advice based on their experiences and encourage their children to put in the effort to learn. Additionally, parents' level of education corresponds to their attitude toward education thus their goals for their children. These factors are all positively correlated with students' academic success. Parents without post-secondary education are less likely to have discussions and create learning opportunities in the household (Idris,2020). Thus parents with more years of education are found to aspire their children to have higher skills, better equipping them for post-secondary education. 
  Another study done in Vietnam used a mean and quantile regression approach to observe the effect of household income on higher levels of education (Vu, 2020). The key finding highlighted that higher levels of education are positively correlated with earning higher income. This was very clear when put into the context of Vietnam as a country. Vietnam was known as one of the poorest countries in the world but in the last few decades has worked its way up to a lower-middle income country. The economy was greatly enhanced by refining the education systems as well as access to tertiary levels (Vu, 2020). This also boosted poverty levels and increased opportunities for children to engage in higher levels of learning. The regression analysis proves that it pays off to pursue higher education as it provides a  positive return for all income groups. Overall, the findings of the study explain that additional years of education results in higher income households as compared to households with fewer years of education who are subject to lower levels of income (Vu, 2020). 
  In this paper, we examine the relationship between one’s parents' level of education, their level of education as well as their income. More specifically we are looking at data from France taken from a European Union Survey in the year 2018. The variables of interest include both the mothers, fathers and childs highest level of education which is ranked on a rating scale from zero through seven. The fourth variable is the child's household net income based on the head of the household.  We are interested in whether there is a correlation between parents' level of education and their child's level of education. Specifically, we are curious as to whether our data has the same trends as the study done in Pakistan, in that parents with high levels of education raise their children to also complete higher education. Furthermore, we are interested in the relationship between one’s level of education and their income. Explicitly, we are keen to understand whether our data and that of the Vietnam study are positively correlated, indicating higher levels of education are analogous to higher income. 
  To better visualize the data, we created figures illustrating different categories of education and where the mother, father and child fit in as well as the childs net household’s income. This was best represented using line plots to analyze the relationship between each variable. Our results indicate that the France data showed similar trends to both Pakistan and Vietnam. Thus we were able to conclude that children with mothers and fathers that have higher levels of education are more likely to themselves engage in higher levels of education. Moreover, it was observed that more years of education are affiliated with higher levels of education. This investigation is captivating as one of the big decisions upon completing secondary school is whether to pursue higher education or enter straight into the workforce. This decision is highly affected by the mother and fathers level of higher education. These two paths yield very different salaries and hence lifestyles therefore highlighting the weight parents' education holds on a child's future finances. 

The remainder of this paper is structured as follows: Section 2 discusses the data, Section 3 presents the results and lastly Section 4 discusses our findings as well as areas of weaknesses and improvements. 


# Data {#sec-data}
I retrieved the data from Open Data Tronto Portal through the library @citeOpenDataToronto. The dataset @redlightcameraannualcharges was used to analyze the intersections with red light cameras across the City of Toronto. After downloading the dataset, I proceeded to the cleaning and analyzing stage with open source statistically programming language R @citeR, using function from tidyverse @citetidyverse, dplyr @citedplyr, ggplot2 @citeggplot2, and kableExtra @citekableExtra.

  The raw dataset contained over 300 intersections across the years 2007 to 2022. I decided to simplify and isolate five intersections across the City of Toronto. Additionally, I decided to analyze years 2012 to 2022 to allow for a ten year analysis.

  After cleaning the data, I generated Table 1, shown below. 
Here I briefly describe the Red Light Data model used to investigate two aspects of the ticket charges. Firstly, it is to explore how the charges change from year to year for individual locations. Secondly, I summed up the charges over the ten years and included a total charges column. This allows for comparison of charges at different intersections in the past ten years. The different locations are interesting as two are located near Lake Ontario, within a kilometer namely Richmond street at Parliament Street and Lake Shore Boulevard at York. On the other hand the other three are in North York, about 9 kilometers from York University, namely Steeles Avenue at Carpenter Road, Steeles Avenue at Hilda Avenue and Albion Road at Silverstone Drive. 

  This model assumes North York is considered as part of the City of Toronto. It's important to note that this data includes only when the driver is charged for running a red light and thus may be missing data from instances where the driver gets away without a ticket. 

```{r}
#| echo: false 
# code not included in doc
#| warning: false 
# no warnings in the doc
#| message: false 
# no messages in the doc
  
### Stimulate data ###
table1 <- cleaned_red_light_data

# rename the columns 
colnames(table1) = c("Location", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022")

# add a column for the sum of charges called Total_Charges  
table1$Total_Charges <- rowSums(sapply(table1[,2:12],function(x)as.numeric(as.character(x))))

# make table cuter
cutie <- kable(table1, format = 'latex', booktable = TRUE, caption = 'Total Number of Charges per Year Based on Location') %>% kable_styling(latex_options = 'scale_down')
cutie

```

# Results

 The outcome of the analysis was to understand how charges for running a red light change every year. I was interested in whether there would be a decrease in charges across all locations because of the $325 fee if you were issued a ticket. I wondered whether proximity to what I consider more downtown Toronto or the CN tower compared to in North York would make a difference as to the number of red light charges given.

```{r}
#| echo: false 
# code not included in doc
#| warning: false 
# no warnings in the doc
#| message: false 
# no messages in the doc

# Want to plot location on x axis and sum charges  on y axis
# creating meaningful axies names, x is intersection and y is Total Charges
bar1 <- table1[,c(1,13)]
ggplot(table1 = bar1$Location) + geom_col(aes(x= table1$Location, y = table1$Total_Charges)) +labs(title = "Graph 1: Location vs. Total Changes", y = "Total Charges", x = "Location") 


```

  Graph 1 includes the total number of charges over the ten years for the five individual locations. I included a bar chart to get a better visual and allow for comparison. Here it can be seen that Richmond Street at Parliament showed the most red light charges in the last ten years, doubling every other intersection. It is followed by Steeles Avenue at Carpenter Road with a difference of over a 1,000 fewer charges. The intersection with the fewest total red light charges was Lake Shore Boulevard at York Street.

# Discussion

  In this paper, I looked at data  concerning the annual number of charges issued to drivers that ran red lights. More specifically, I cleaned the dataset to investigate five different intersections in the City of Toronto across a ten year period. I created a table, namely table 1 in the data section illustrating the annual charges for each location. This allowed a visual aid to illustrate how there is an overall trend of an increase in charges in four out of five intersections. The exception is Albion Road at Silverstone Drive which shows a slight decrease in charges over the ten years however the data on the most recent five years does illustrate a surplus of charges. Additionally, four out of five of the intersections do show an increase in the last five years aside from Richmond Street and Parliament Street. It is interesting that Richmond Street at Parliament Street showed steady levels for the first few years, followed by a large spike in the number of charges and in the recent five years has declined. From the table and graphs, we can take away a general increase in charges; however it is important to note that this depiction is from a small sample of five intersections in different areas in the City of Toronto. 

  This paper was eye opening in that despite a $325 charge for running a red light, many drivers still run lights and in fact there has been an overall increase in drivers who are issued tickets. Although I did not deeply investigate the type of intersection, I think it is an important factor to consider in terms of understanding which intersections drivers are more likely to run a red light. Additionally, there are many other factors that need to be considered when interested in the likelihood of drivers getting charged. It would also be interesting to see data on drivers that are not charged for running red lights. However some data may be hard to come by and or there may not be enough data to make a good analysis.  This dataset however, contained over 300 intersections however only 80 had data for ten consecutive years. I should note that the data only includes drivers that did get charged for running the light and thus there could be data missing for those drivers that did not get charged. 

  I learned that consecutive and consistent data is important to create justifiable assumptions. Moreover, an important aspect of working with data is deciding what is important and picking a focus. One weakness is the number of intersections I looked at. I think it would have been more meaningful in allowing for a larger sample. 

  Future work could look at intersections close to one another to get an idea about which types of intersections drivers are more likely to run red lights. For example, it would be interesting and relevant to pick intersections around the University of Toronto and analyze the number of charges issued. This would impact student safety and thus perhaps the data would show a decrease in charges around the area. However having another area to compare the data to is equally important.  More consistent data is needed  across consecutive periods of time to really ensure correct assumptions are made and portray accurate data. THis also improves others' trust in the data. 

\newpage


# References


